### 2020-02-09 doc/icfp20/counterpoint.tex
explain FirstSpecies record type

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which starts with a tonic and consists only of whole notes.
Figure~\ref{fig:frog} is an example of first species counterpoint,
where the lower line stands for the cantus firmus and the upper line
represents the counterpoint\footnote{The cantus firmus is part of a
  German song called ``Froschgesang'' (Frog's song).}:

Among different options of how to encode such music, we adopt
the following strategy.
First, we represent each bar as a pitch-interval pair \texttt{(p ,  i)},
where \texttt{p} is a note in the cantus firmus, and \texttt{i} is the
distance between \texttt{p} and the corresponding counterpoint note.
Second, we keep the first and last bars as separate elements, and
represent the middle bars (often referred to as the \emph{main body})
as a list of pitch-interval pairs.
For instance, the main body of the above example is represented as
follows:

@@

cfcp1 : List PitchInterval
cfcp1 = (d 5 , maj6) :: (e 5 , min6) :: (f 5 , maj3) :: (e 5 , min3) :: (d 5 , maj6) :: []
\end{alltt}

The reason we separate different parts of the music is that they are
subject to different rules.
In the reminder of this section, we elaborate these rules one by one.

\paragraph{Beginning}
The beginning of the music should express perfection.
Therefore, the first interval must be a unison, a 5th, or an 8th.
In our formalization, we implement this rule as a function
\texttt{checkBeginning}, which reports an error \texttt{not158 i}
when the first interval \texttt{i} is an invalid one.

@@

\paragraph{Intervals}
The main body of the music should maintain harmonical consonance
and independence of melodic lines.
This means, all constituent intervals must be chosen from the 3rd,
5th, 6th, and 8th.
We encode this rule as the \texttt{checkIntervals} function,
which returns a list of errors correponding to the occurrences of
dissonant intervals and unisons.

@@

\paragraph{Motion}
The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As shown in Figure~\ref{fig:motion}, there are four kinds of motion,
and among them, parallel and similar motion makes it harder to
distinguish between two voices, especially when the second interval
is a perfect one.
To avoid this problematic motion, we define the \texttt{checkMotion}
function, which lists all invalid motion approaching a perfect interval.

@@

\paragraph{Ending}
The ending of the music should express relaxation.
This means, the last interval must be chosen from the unison and
the 8th, which are both considered very stable.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution.
We encode these rules as the \texttt{checkEnding} function;
note that it returns the \texttt{tooShort} error when the music does
not have enough number of bars to form a valid ending.

@@

>>>>

First species counterpoint is the simplest variant of counterpoint.
In first species, we set one note against each note in the cantus firmus,
which starts with a tonic and consists only of whole notes.
Figure~\ref{fig:frog} is an example of first species counterpoint.
The lower line is the cantus firmus, which we take from a famous
German song called ``Froschgesang'' (Frog's song).
The upper line is the counterpoint composed by the second author.

\subsubsection{Representing Music}

In our implementation, we represent each bar as a pitch-interval
pair \texttt{(p ,  i)}, where \texttt{p} is a note in the cantus firmus,
and \texttt{i} is the distance between \texttt{p} and the corresponding
counterpoint note.
We then represent the sequence of bars as a list of pitch-interval pairs,
but with one proviso: we separate the first and last bars from the
middle bars.
Therefore, the Frog's song in Figure~\ref{fig:frog} is represented as a
compound of the following three elements:

@@

first : PitchInterval
first = (c 5 , per8)

middle : List PitchInterval
middle = (d 5 , maj6) :: (e 5 , min6) :: (f 5 , maj3) :: (e 5 , min3) :: (d 5 , maj6) :: []

last : PitchInterval
last = (c 5 , per8)
\end{alltt}

\subsubsection{Representing Rules}

The reason why we separate different parts of music is that
they are subject to different rules.
Below, we elaborate these rules one by one.

\paragraph{Beginning}

The beginning of the music should express perfection.
Therefore, the first interval must be a unison, a 5th, or an 8th.
In our formalization, we implement this rule as a function
\texttt{checkBeginning}, which reports an error \texttt{not158 i}
when the first interval \texttt{i} is an invalid one.

@@

\paragraph{Intervals}

The middle bars of the music should maintain harmonical
consonance and independence of melodic lines.
This means, all constituent intervals must be chosen from the 3rd,
5th, 6th, and 8th.
We encode this rule as the \texttt{checkIntervals} function,
which returns a list of errors correponding to the occurrences of
dissonant intervals and unisons.

@@

\paragraph{Motion}

The independence of melodic lines is also affected by \emph{motion},
i.e., the way one interval moves to another interval.
As shown in Figure~\ref{fig:motion}, there are four kinds of motion,
and among them, parallel and similar motion makes it harder to
distinguish between two voices, especially when the second interval
is a perfect one.
To avoid this problematic motion, we define the \texttt{checkMotion}
function, which lists all invalid motion approaching a perfect interval.

@@

\paragraph{Ending}

The ending of the music should express relaxation.
This means, the last interval must be chosen from the unison and
the 8th, which are both considered very stable.
The last interval should also be approached by a \emph{cadence},
a progression that gives rise to a sense of resolution.
We encode these rules as the \texttt{checkEnding} function;
note that it returns the \texttt{tooShort} error when the music does
not have enough number of bars to form a valid ending.

@@

\subsubsection{Putting Things Together}

Using the encoding of music and rules we have seen so far,
we define \texttt{FirstSpecies}, a record type inhabited by correct
counterpoint.
The first three fields of this record type represent the first, middle,
and last bars, respectively.
The last four fields stand for the proofs that the music satisfies
all the required properties, that is:

\begin{itemize}
  \item The beginning is a perfect consonance
  \item The middle bars have no dissonant interval or unison
  \item The middle bars have no perfect interval approached by
    parallel or similar motion
  \item The ending is a unison or an octave approached by a cadence
\end{itemize}

\begin{alltt}
record FirstSpecies : Set where
  constructor firstSpecies
  field
    firstBar    : PitchInterval
    middleBars  : List PitchInterval
    lastBar     : PitchInterval
    beginningOk : checkBeginning firstBar \(\equiv\) nothing
    intervalsOk : checkIntervals middleBars \(\equiv\) []
    motionOk    : checkMotion middleBars \(\equiv\) []
    endingOk    : checkEnding middleBars lastBar \(\equiv\) nothing
\end{alltt}

With this record type, we can show that the counterpoint composed
for the Frog's song is correct, since the music inhabits the
\texttt{FIrstSpecies} type.

\begin{alltt}
fs : FirstSpecies
fs = firstSpecies first1 middle1 last1 refl refl refl refl
\end{alltt}
